---
title: "Tutorial: Feature selection with the FBED algorithm - method: eBIC"
author:
- name: Kleio - Maria Verrou
  affiliation:
  - Medicine Department, University of Crete, Greece
  - Mens ex Machina Group, Computer Science Department, University of Crete, Greece

- name: Michail Tsagris
  affiliation: Mens ex Machina Group, Computer Science Department, University of Crete, Greece
  email: mtsagris@csd.uoc.gr

output:
  BiocStyle::html_document2:
    toc_float: true
  BiocStyle::pdf_document2: default

vignette: |
  %\VignetteIndexEntry{Feature selection with FBED}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<br>
<br>
<img src = 'http://bit.ly/mxm_FBED_image_png' width="500" height="1000" align="middle" />
<br>
<br> 


# Introduction
  
  
The MXM R Package, short for the latin 'Mens ex Machina' ( Mind from the Machine ), is a collection of utility functions for feature selection, cross validation and Bayesian Networks. MXM offers many feature selection algorithms focused on providing one or more minimal feature subsets, refered also as variable signatures, that can be used to improve the performance of downstream analysis tasks such as regression and classification, by excluding irrelevant and redundant variables.
  
In this tutorial we will learn how to use the *Forward Backward Early Dropping (FBED)* algorithm. The algorithm is a variation of the usual forward selection. At every step, the most significant variable enters the selected variables set. In addition, only the significant variables stay and are further examined. The non significant ones are dropped. This goes until no variable can enter the set. The user has the option to redo this step 1 or more times (the argument K). In the end, a backward selection is performed to remove falsely selected variables.
  
For simplicity, in this tutorial,  we will use a dataset referred as **"The Wine Dataset"**. 

# Loading Data
**The Wine Dataset** contains the results of a chemical analysis of wines grown in a specific area of Italy. Three types of wine are represented in the 178 samples, with the results of 13 chemical analyses recorded for each sample. Note that the "Type" variable was transformed into a categorical variable.


So, first of all, for this tutorial analysis, we are loading the 'MXM' library and 'dplyr' library for handling easier the dataset. The 'hash' library used also be imported, if an analysis is going to be applied more than once (we will explain this in the **TIPS** session).  

```{r, warning =  FALSE, message = FALSE }

### ~ ~ ~ Load Packages ~ ~ ~ ###
library(MXM) 
library(dplyr)
```

And on a next step we are downloading and opening the dataset, defining also the column names.

```{r}

### ~ ~ ~ Load The Dataset ~ ~ ~ ###
wine.url <- "ftp://ftp.ics.uci.edu/pub/machine-learning-databases/wine/wine.data"
wine <- read.csv(wine.url,
                 check.names = FALSE,
                 header = FALSE) 
head(wine)
str(wine)
colnames(wine) <- c('Type', 'Alcohol', 'Malic', 'Ash', 
                    'Alcalinity', 'Magnesium', 'Phenols', 
                    'Flavanoids', 'Nonflavanoids',
                    'Proanthocyanins', 'Color', 'Hue', 
                    'Dilution', 'Proline')
```


# FBED for Continuous

For this tutorial example, we are going to apply the FBED algorithm on the above dataset, using as data and as target variables only continuous variables.   


##  Running FBED for 1st time


### Selecting Appropriate Conditional Independence Test   


The selection of the appropriate conditional independence test is a crucial decision for the validity and success of downstream statistical analysis and machine learning tasks. Currently the __ ` MXM R package`__  supports numerous tests for different combinations of __target__ ( _dependent_ ) and __predictor__ ( _independent_ ) variables. A detailed summary table to guide you through the selection of the most suitable test can be found in __MXM's__ reference manual (p.21 _"CondInditional independence tests"_ ) here: <https://CRAN.R-project.org/package=MXM>. 
In our example we will use the __`MXMX::fbed.reg()`__, which is the implementation of the FBED algorithm and since we are going to examine only continuous variables, we will use the Fisher's Independence Test.


### Creating Data & Target Matrices


`dataset` - A numeric matrix ( or a _data.frame_ in case of categorical predictors), 
containing the variables for performing the test. The rows should refer to the different samples and columns to the features. For the purposes of this example analysis, we are going to use only the continuous variables, therefore we are removing the "Type" variable from the dataset. Furthermore, we are removing the "Nonflavanoids" variable, because we will use it as target.


```{r}
### ~ ~ ~ Removing The Categorical ('Type') and The Target ('Nonflavanoids') Variables ~ ~ ~ ###

wine_dataset <- dplyr::select(wine,
                              -contains("Type"),
                              -contains("Nonflavanoids")) 
head(wine_dataset)
```


`target` -  The class variable including the values of the target variable. We should provide either a string, an integer, a numeric value, a vector, a factor, an ordered factor or a Surv object.
For the purposes of this example analysis, we are going to use as the dependent variable Nonflavanoids. 

```{r}
wine_target <- wine$Nonflavanoids
head(wine_target)
```


###  Function's Arguments

This is the first time that we are running the algorithm, so we are going to explain what each **Argument** refers to:
  
  `y` : The class variable. Provide either a string, an integer, a numeric value, a vector, a factor, an ordered factor or a Surv object. As explained above, this will be the dependent variable. If the target is a single integer value or a string, it has to corresponds to the column number or to the name of the target feature in the dataset. *Here* we choose wine$Nonflavanoids

`x` : The dataset. Provide either a data frame or a matrix. If the dataset (predictor variables) contains missing (NA) values, they will automatically be replaced by the current variable (column) mean value with an appropriate warning to the user after the execution. *Here* we choose the whole wine dataset, except from the Type (categorical) and Nonflavanoids (target) variables.
  
`test` : The conditional independence test to use. Default value is NULL. *Here* since our dataset includes only continuous features (*remember*: Categorical variable 'Type' was removed) and our dependent variable is also continuous, we choose 'testIndFisher'.
For more information, about which test to use, please visit :  <https://www.rdocumentation.org/packages/MXM/versions/0.9.7/topics/CondInditional%20independence%20tests>. 
  
`alpha` : Threshold (suitable values in [0,1]) for the significance of the p-values.The default value is 0.05. *Here* we choose the default value 0.05.  
  
`wei` : A vector of weights to be used for weighted regression. The default value is NULL. It is not suggested when robust is set to TRUE. If you want to use the "testIndBinom", then supply the successes in the y and the trials here. *Here* we choose the default value NULL
  
`K` : How many times should the process be repeated? The default value is 0. *Here* we choose 3.
  
`method` : Do you want the likelihood ratio test to be performed ("LR" is the default value) or perform the selection using the "eBIC" criterion (BIC is a special case). *Here* we choose BIC.
  
`gam` : In case the method is chosen to be "eBIC" one can also specify the gamma parameter. The default value is "NULL", so that the value is automatically calculated. *Here*, although we choose BIC as selection criterion, we do not choose any gamma parameter.
  
`backward` : After the Forward Early Dropping phase, the algorithm proceeds with the usual Backward Selection phase. The default value is set to TRUE. It is advised to perform this step as maybe some variables are false positives, they were wrongly selected. The backward phase using likelihood ratio test and eBIc are two different functions and can be called directly by the user. So, if you want for example to perform a backward regression with a different threshold value, just use these two functions separately. *Here* we set the backward argument as TRUE 




```{r}
### ~ ~ ~ Running FBED For First Time ~ ~ ~ ###
fbed_default_1st <- MXM::fbed.reg(target       = wine_target,
                                 dataset        = wine_dataset, 
                                 test     = "testIndFisher", 
                                 threshold    = 0.05,
                                 wei      = NULL,
                                 K        = 10,
                                 method   = "eBIC",
                                 gam      = NULL,
                                 backward = TRUE)
```


So, the algorithm run...
Let's see what information we can take out of it.  


### Output


The main purpose of running FBED algorithm is to see which variables should be selected as important. The indices of those variables are stored in `res`. Furthermore, in this matrix we see their test statistic and the associated p-value.  

```{r}
fbed_default_1st$res
SelectedVars_names<-colnames(wine_dataset[fbed_default_1st$res[,1]])
SelectedVars_names
```

Here, we see which are the important features. Yes yes, as you see in the second column of `res` matrix, they are sorted... don't worry!

  
And as you may imagine, you may also retrieve the information about their scores. They are all (sorted) in the second column. 
``` {r}
fbed_default_1st$res[,2]
```
  
Perfect! But we see that the function returned an object called `info`. What is this?  

```{r}
fbed_default_1st$info
```
The `info` matrix describes the number of variables and the number of tests performed (or models fitted) at each round (remember this value of `K` that in this example we set equal to 10. *Here* it did not reach K=10, because there were no difference after the 3rd run, so the algorithm stopped running earlier). Well, this refers to the forward phase only and for each `K` the number of selected variables is returned together with the number of tests performed.
  
So, if the information about the forward step is appended in the `info` matrix, where can we find information about the backward phase?
 
```{r}
fbed_default_1st$back.rem 
``` 
  
By calling the `back.rem`, the variables that have been removed in the backward phase are returned. 
In case we are interested in the number of models that were fitted in the backward phase, all we have to do is to look for the `back.n.tests` variable.

```{r}
fbed_default_1st$back.n.tests 
```  

And how quick has all this happened? 

```{r}
fbed_default_1st$runtime 
```  


## Running FBED for Categorical 


### Selecting Appropriate Conditional Independence Test

Since the variable is categorical - and more specific it is a factor with more than two levels (unordered) -and the features are continuous, according to __MXM's__ reference manual (p.21 _"CondInditional independence tests"_ ) here: <https://CRAN.R-project.org/package=MXM>, we should use the Multinomial logistic regression ( 'testIndMultinom' ).

### Creating Data & Target Matrices

In this step, we keep the whole dataset, in order to show how to use the algorithm also without subtracting the initial matrix.

```{r}
### ~ ~ ~ Taking The Whole Dataset ~ ~ ~ ###
wine_dataset <- wine
head(wine_dataset)
```  

We will not create a different matrix for the target. As mentioned above, we are going to use the Type variable, but please... be patient...

### Setting the Arguments 

```{r, message=FALSE}
### ~ ~ ~ Running FBED For Categorical Variable ~ ~ ~ ###
wine[, 1] <- as.factor(wine[, 1])
fbed_default_2nd <- MXM::fbed.reg(target       = wine_target,
                                 dataset        = wine_dataset, 
                                 test     = "testIndReg", 
                                 threshold    = 0.05,
                                 wei      = NULL,
                                 K        = 10,
                                 method   = "eBIC",
                                 gam      = NULL,
                                 backward = TRUE) 

```  

So, the algorithm run once again...
Let's see what information we can take out of it.

### Output

The main purpose of running FBED algorithm is to see which variables should be selected as  important. The indices of those variables are stored in `selectedVars`.

```{r}
fbed_default_2nd$res
SelectedVars_names<-colnames(wine_dataset[fbed_default_2nd$res[,1]])
SelectedVars_names
```  
  
Again we see the scores in the second column.
``` {r}
fbed_default_2nd$res[,2]
```
  
What was stored this time in the `info` matrix?

```{r}
fbed_default_2nd$info
```
The `info` matrix describes again the number of variables and the number of tests performed (or models fitted) at each round (remember this value of `K` that in this example we set equal to 10. *Here* again it did not reach K=10, because there were no difference after the 2nd run, so the algorithm stopped earlier). Well, this refers to the forward phase only and for each `K` the number of selected variables is returned together with the number of tests performed.
  
And now let us inspect the backward phase
 
```{r}
fbed_default_2nd$back.rem 
fbed_default_2nd$back.n.tests 
```  

And how quick has all this happened? 

```{r}
fbed_default_2nd$runtime 
```    




# Conclusion


>Now you are ready to run your own analysis using MXM::FBED algorithm!   
>Thank you for your attention.    
>Hope that you found this tutorial helpful.    



# Session Info {.unnumbered}  
All analyses have been applied on:

```{r}
sessionInfo()
```



# References {.unnumbered}


Borboudakis G. and Tsamardinos I. (2017). Forward-Backward Selection with Early Dropping. https://arxiv.org/pdf/1705.10770.pdf


